Fitting the Undetected: Light-curve Feature Estimation in the Presence of Non-detections

### 2012-04-10 tex/ms.tex
small typo corrected
add two hyperparameters, small Latex tweaks, drop and ignore pdf file

Hence, our initial model consists of the parameter vectors $\{f^*_1,...,f^*_N\}$, $\{b_1,...,b_N\}$ and $\{\sigma^2_1,...,\sigma^2_N\}$, and the model parameters
\begin{eqnarray}\displaystyle
\theta &\equiv& (\omega, A_0, \{A_k, B_k\}_{k=1}^K, s_\mu^2, \cdots) \quad ,
\end{eqnarray}
along with prior information about observation times, $\{t_1,...,t_N\}$ and other prior assumptions \joey{we should be more explicit here}, which we make explicit by creating the prior information set
\begin{eqnarray}\displaystyle
I &\equiv& (\{t_i\}_{i=1}^N, \mbox{assumptions}) \quad .
\end{eqnarray}
Our goal is to write down the form of the likelihood of the data, $D$, given $\theta$ and $I$.  Then, for each light curve we can search for the vector $\theta$ that maximizes the data likelihood, and use those maximum likelihood estimates for downstream astrophysical inference.

@@

We model the observed flux, $\fobs$, as a Gaussian distribution with variance that has both measurement ($\sigma^2_i$) and model ($s_{\mu}$) contributions.  In the case that a detection is made, we require that the observed brightness be greater than the brightness of the detection limit ($\fobs = \ftrue \ge b_i$) while in the case that no detection is made, we require that the brightness (if it could be measured) be less than the detection limit ($\ftrue < b_i$).  To derive the likelihood of the observed $\fobs$, given $\sigma^2_i$, $\theta$ and $I$, we must integrate over the prior distribution of the unknown $b_i$,
\begin{eqnarray}\displaystyle
p(\fobs |\sigma^2_i,\theta,I) &=& \left\{\begin{array}{ll}
  N(\ftrue | \mu_i,  \sigma^2_i + s_{\mu}^2)\,  \int_0^{\ftrue} p(b_i | \theta)\, \dd b_i & \mbox{if $\fobs \ne \texttt{NA}$} \\
  \int_{0}^{\infty} \int_{0}^{b_i} N(\ftrue | \mu_i, \sigma^2_i + s_{\mu}^2)\, p(b_i | \theta)\, \dd \ftrue\, \dd b_i & \mbox{if $\fobs = \texttt{NA}$} \\
\end{array}\right.\label{eq:mlik}
\\
p(b_i|\theta) &=& N(b_i|B,V_B)
\label{eq:bprior}
\\
\theta &\equiv& (\omega, A_0, \{A_k, B_k\}_{k=1}^K, s_\mu^2, B, V_B, \cdots) \quad ,
\end{eqnarray}
where we have introduced the hyperparameters $B$ and $V_B$ for the Gaussian prior distribution of $b_i$.  In Equation \ref{eq:mlik}, in the epochs for which a detection was made ($\fobs \ne \texttt{NA}$), we marginalize over the unknown $b_i$ from $0$ (low brightness limit) to $\ftrue$, enforcing that the detection limit be fainter than the observed brightness.  Likewise, in the epochs for which no detection was made ($\fobs = \texttt{NA}$), we integrate the joint $(\ftrue, b_i)$ likelihood over all possible values of $b_i$ and over the unknown $\ftrue$ from $0$ to $b_i$, ensuring that the brightness (if it were able to be observed) be fainter than the detection limit.

In the above, we have assumed no extra information on each of the $b_i$ values besides the knowledge of whether a detection was made on that epoch.  Hence, we draw, in Equation \ref{eq:bprior}, each $b_i$ value from a global prior distribution which is the same at all epochs.  If instead, we are given an estimate of $b_i$ plus its error distribution for each epoch (which, in principle can be inferred from the raw telescope images), we can replace Equation \ref{eq:bprior} with a different distribution per epoch.  In the case that the $b_i$ are assumed to be completely known (without error), the data likelihood of $\fobs$ becomes
\begin{eqnarray}\displaystyle
p(\fobs |\sigma^2_i,\theta,I, b_i) &=& \left\{\begin{array}{ll}
  N(\ftrue | \mu_i,  \sigma^2_i + s_{\mu}^2)\,  I(\ftrue \ge b_i) & \mbox{if $\fobs \ne \texttt{NA}$} \\
 \int_{0}^{b_i} N(\ftrue | \mu_i, \sigma_i^2 + s_{\mu}^2)\, \dd \ftrue & \mbox{if $\fobs = \texttt{NA}$} \\
\end{array}\right.\label{eq:mlik_s}
\end{eqnarray}
where the boolean indictor function, $I(\ftrue \ge b_i)$, is 1 if $\ftrue \ge b_i$ and 0 otherwise.

Next, we model the reported variance, $\sobs$, on the uncertainty of the magnitude measurement.  Instead of assuming that $\sobs$ is a perfect, error-free measurement, we probabilistically connect it to the true uncertainty variance, $\sigma^2_i$ through a Gamma likelihood, which is defined to take on non-negative values.  Our likelihood of observed $\sobs$, given $\sigma^2_i$, $\theta$ and $I$, is
\begin{eqnarray}\displaystyle
p(\sobs | \fobs, \sigma^2_i, \theta,I) &=& \left\{\begin{array}{ll}
  \Gamma (\sobs | \sigma^2_i, V_{\sigma} ) & \mbox{if $\fobs \ne \texttt{NA}$} \\
 1& \mbox{if $\fobs = \texttt{NA}$}
\end{array}\right.\label{eq:slik}
\end{eqnarray}
where $\Gamma(x | m, V)$ is the standard Gamma distribution with mean $m$ and variance $V$, 
and we have added a model parameter $V_\sigma$ to represent the
variance in the distribution of reported uncertainties given the true
uncertainty.  (Typically, the Gamma distribution is parameterized by parameters $(\alpha, \beta)$, where the mean is $\alpha\beta$ and the variance is $\alpha\beta^2$.)  The  purpose of 
 including the likelihood in Equation \ref{eq:slik} to the model, in practice, to keep the $\sigma^2_i$ from
drifting very far away from the $s^2_i$, as set by the hyperparameter
$V_\sigma$.

Putting it all together, we can write down the likelihood of the data, $D_i$, for the parameters $\theta$, on a single epoch, as
\begin{eqnarray}\displaystyle
p(D_i|\theta,I) &=& \int_0^{\infty} p(\fobs |\sigma^2_i,\theta,I)\, p(\sobs | \fobs, \sigma^2_i,\theta,I)\, p(\sigma^2_i | \theta, I)\,\dd \sigma^2_i
\\
p(\sigma^2_i|\theta, I) &\propto& \sigma_i^{-1}
\\
\theta &\equiv& (\omega, A_0, \{A_k, B_k\}_{k=1}^K, s_\mu^2, B, V_B, V_{\sigma}) \quad ,
\end{eqnarray}
where we have inserted the expressions from Equations (\ref{eq:mlik}) and (\ref{eq:slik}), and integrated out the nuisance parameter, $\sigma^2_i$.  We have assumed a Jeffrey's prior on $\sigma^2_i$, which is non-informative and invariant to reparametrization of the variance.  We could alternatively use an inverse-gamma prior, which takes two hyperparameters (and is conjugate in the case of a normal likelihood with unknown variance). % We have also assumed that $\fobs$ and $\sobs$ are conditionally independent given $\sigma^2_i$, $\theta$, and $I$.  This is a reasonable assumption in the case that a detection is made at epoch $t_i$.

@@

To simulate faint Mira stars from ASAS 235627-4947.2, we use the following procedure:
\begin{enumerate}
\item Convert the observed magnitudes, $m_i$, and errors, $s_{m,i}$ to fluxes, $f_i$ and flux errors, $s_{f,i}$ (we assume a V-band zero-point of $3.67\times10^{-9}$ erg/s/cm$^2$/\AA)
\item For a given flux dimming parameter, $d$, sample the new fluxes, $\tilde{f}_i$, from a Gaussian distribution centered around $f_i / d$ with standard deviation sampled from the empirical ASAS distribution of $s_{f,i} | f_i$. \joey{I found that relationship to be linear with a slope $\approx 1$, which is probably BS since the ASAS mag errors are all crap.  So I added a reference value to each flux error to ensure a mag limit $\approx$ 14.5 mag.}
\item Fluxes that are not at least 5$\sigma$ above zero are denoted as non-detections and their flux estimates (and errors) are censored.
\end{enumerate}
Folded light curves of ASAS 235627-4947.2, dimmed by four different values of $d$, are plotted in Figure \ref{fig:mirasim}.

>>>>

Hence, our initial model consists of the parameter vectors $\{f^*_1,...,f^*_N\}$, $\{b_1,...,b_N\}$ and $\{\sigma^2_1,...,\sigma^2_N\}$, and the model parameters
\begin{eqnarray}\displaystyle
\theta &\equiv& (\omega, A_0, \{A_k, B_k\}_{k=1}^K, s^2_\mu, \cdots) \quad ,
\end{eqnarray}
along with prior information about observation times, $\{t_1,...,t_N\}$ and other prior assumptions \joey{we should be more explicit here}, which we make explicit by creating the prior information set
\begin{eqnarray}\displaystyle
I &\equiv& (\{t_i\}_{i=1}^N, \mbox{assumptions}) \quad .
\end{eqnarray}
Our goal is to write down the form of the likelihood of the data, $D$, given $\theta$ and $I$.  Then, for each light curve we can search for the vector $\theta$ that maximizes the data likelihood, and use those maximum likelihood estimates for downstream astrophysical inference.

@@

We model the observed flux, $\fobs$, as a Gaussian distribution with variance that has both measurement ($\sigma^2_i$) and model ($s^2_{\mu}$) contributions.  In the case that a detection is made, we require that the observed brightness be greater than the brightness of the detection limit ($\fobs = \ftrue \ge b_i$) while in the case that no detection is made, we require that the brightness (if it could be measured) be less than the detection limit ($\ftrue < b_i$).  To derive the likelihood of the observed $\fobs$, given $\sigma^2_i$, $\theta$ and $I$, we must integrate over the prior distribution of the unknown $b_i$,
\begin{eqnarray}\displaystyle
p(\fobs |\sigma^2_i,\theta,I) &=& \left\{\begin{array}{ll}
  N(\ftrue | \mu_i,  \sigma^2_i + s^2_{\mu})\,  \int_0^{\ftrue} p(b_i | \theta)\, \dd b_i & \mbox{if $\fobs \ne \texttt{NA}$} \\
  \int_{0}^{\infty} \int_{0}^{b_i} N(\ftrue | \mu_i, \sigma^2_i + s^2_{\mu})\, p(b_i | \theta)\, \dd \ftrue\, \dd b_i & \mbox{if $\fobs = \texttt{NA}$} \\
\end{array}\right.\label{eq:mlik}
\\
p(b_i|\theta) &=& N(b_i|B,V_B)
\label{eq:bprior}
\\
\theta &\equiv& (\omega, A_0, \{A_k, B_k\}_{k=1}^K, s^2_\mu, B, V_B, \cdots) \quad ,
\end{eqnarray}
where we have introduced the hyperparameters $B$ and $V_B$ for the Gaussian prior distribution of $b_i$.  In Equation \ref{eq:mlik}, in the epochs for which a detection was made ($\fobs \ne \texttt{NA}$), we marginalize over the unknown $b_i$ from $0$ (low brightness limit) to $\ftrue$, enforcing that the detection limit be fainter than the observed brightness.  Likewise, in the epochs for which no detection was made ($\fobs = \texttt{NA}$), we integrate the joint $(\ftrue, b_i)$ likelihood over all possible values of $b_i$ and over the unknown $\ftrue$ from $0$ to $b_i$, ensuring that the brightness (if it were able to be observed) be fainter than the detection limit.

In the above, we have assumed no extra information on each of the $b_i$ values besides the knowledge of whether a detection was made on that epoch.  Hence, we draw, in Equation \ref{eq:bprior}, each $b_i$ value from a global prior distribution which is the same at all epochs.  If instead, we are given an estimate of $b_i$ plus its error distribution for each epoch (which, in principle can be inferred from the raw telescope images), we can replace Equation \ref{eq:bprior} with a different distribution per epoch.  In the case that the $b_i$ are assumed to be completely known (without error), the data likelihood of $\fobs$ becomes
\begin{eqnarray}\displaystyle
p(\fobs |\sigma^2_i,\theta,I, b_i) &=& \left\{\begin{array}{ll}
  N(\fobs | \mu_i,  \sigma^2_i + s^2_{\mu})\,  I(\fobs \ge b_i) & \mbox{if $\fobs \ne \texttt{NA}$} \\
 \int_{0}^{b_i} N(\ftrue | \mu_i, \sigma_i^2 + s^2_{\mu})\, \dd \ftrue & \mbox{if $\fobs = \texttt{NA}$} \\
\end{array}\right.\label{eq:mlik_s}
\quad ,
\end{eqnarray}
where the boolean indictor function, $I(\ftrue \ge b_i)$, is 1 if $\ftrue \ge b_i$ and 0 otherwise.

Next, we model the reported variance, $\sobs$, on the uncertainty of the magnitude measurement.  Instead of assuming that $\sobs$ is a perfect, error-free measurement, we probabilistically connect it to the true uncertainty variance, $\sigma^2_i$ through a Gamma likelihood, which is defined to take on non-negative values.  Our likelihood of observed $\sobs$, given $\sigma^2_i$, $\theta$ and $I$, is
\begin{eqnarray}\displaystyle
p(\sobs | \fobs, \sigma^2_i, \theta,I) &=& \left\{\begin{array}{ll}
  \Gamma (\sobs | \sigma^2_i, V_{\sigma} ) & \mbox{if $\fobs \ne \texttt{NA}$} \\
 1& \mbox{if $\fobs = \texttt{NA}$}
\end{array}\right.\label{eq:slik}
\quad ,
\end{eqnarray}
where $\Gamma(x | m, V)$ is the standard Gamma distribution with mean $m$ and variance $V$, 
and we have added a model parameter $V_\sigma$ to represent the
variance in the distribution of reported uncertainties given the true
uncertainty.  (Typically, the Gamma distribution is parameterized by parameters $(\alpha, \beta)$, where the mean is $\alpha\beta$ and the variance is $\alpha\beta^2$.)  The  purpose of 
 including the likelihood in Equation \ref{eq:slik} to the model, in practice, to keep the $\sigma^2_i$ from
drifting very far away from the $s^2_i$, as set by the hyperparameter
$V_\sigma$.

Putting it all together, we can write down the likelihood of the data, $D_i$, for the parameters $\theta$, on a single epoch, as
\begin{eqnarray}\displaystyle
p(D_i|\theta,I) &=& \int_0^{\infty} p(\fobs |\sigma^2_i,\theta,I)\, p(\sobs | \fobs, \sigma^2_i,\theta,I)\, p(\sigma^2_i | \theta, I)\,\dd \sigma^2_i
\\
p(\sigma^2_i|\theta, I) &=& \Gamma(\sigma^2_i | S, V_S)
\\
\theta &\equiv& (\omega, A_0, \{A_k, B_k\}_{k=1}^K, s^2_\mu, B, V_B, V_{\sigma}, S, V_S) \quad ,
\end{eqnarray}
where we have inserted the expressions from Equations (\ref{eq:mlik}) and (\ref{eq:slik}), we have integrated out the nuisance parameter $\sigma^2_i$, again the $\Gamma$ is the Gamma function with specifid mean and variance, and we have two more hyperparameters.

@@

To simulate faint Mira stars from ASAS 235627-4947.2, we use the following procedure:
\begin{enumerate}
\item Convert the observed magnitudes $m_i$ and uncertainty variances $s^2_{m,i}$ to fluxes $f_i$ and flux uncertainty variances $s^2_{f,i}$ (we assume a V-band zero-point of $3.67\times10^{-9}$ erg/s/cm$^2$/\AA)
\item For a given flux dimming parameter, $d$, sample the new fluxes, $\tilde{f}_i$, from a Gaussian distribution centered around $f_i / d$ with standard deviation sampled from the empirical ASAS distribution of $s^2_{f,i} | f_i$. \joey{I found that relationship to be linear with a slope $\approx 1$, which is probably BS since the ASAS mag errors are all crap.  So I added a reference value to each flux error to ensure a mag limit $\approx$ 14.5 mag.}
\item Fluxes that are not at least 5$\sigma$ above zero are denoted as non-detections and their flux estimates (and errors) are censored.
\end{enumerate}
Folded light curves of ASAS 235627-4947.2, dimmed by four different values of $d$, are plotted in Figure \ref{fig:mirasim}.
